---
title: "Modeling and CEA of malaria vaccine in Kenya"
author: "Catherine Wenger"
date: "2023-12-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggplot2)
```

```{r}
#preset parameters for all models 
#prevalence
pE=0.2
pI=0.2
pR=0.2

#pop size models 1&2
n1=4709493
n2=9418986
n3=37324930
```

No Vaccine:
NOTE: I think the ggplot is pulling the wrong compartment data since its seir, not dsdedidr
```{r}
############# Model Equations ######################

my_model <- function(times,yinit,pars){
	
	with(as.list(c(yinit,pars)), {
	
		N = S + E + I + R
		dS_nv <- w*R-beta*S
		dE_nv <- beta*S - lam*E
		dI_nv <- lam*E - gamma*I
		dR_nv <- gamma*I - w*R

	return(list(c(dS_nv, dE_nv, dI_nv, dR_nv)))})
	}

################## Setting Parameters ##################
#birth rate
b=0.00000575
#transmission parameter
beta=0.0655
#rate of becoming infectious
lam=0.07143
#rate of recovery
gamma=0.005+(0.5*0.0165)
#rate of waning immunity
w=0.01095
#death rates
mu1=0.00005447
mu2=0.00005447
mu3=0.05
mu4=0.00005447
  
my_pars <- cbind(beta,lam,gamma,w) 
  #cbind - combines data-frames, combines columns and rows respectively

############### Setting Time Frame ####################

start_time = 0 # start date (years)
end_time = 50 # end date (years)
my_times <- seq(start_time, end_time, by = 0.25) 
    # gives a sequence from start to end
											 # in increments of 1 month

############ Setting Initial Conditions ###############

# initial starting values
my_yinit <- c(S = n1*(1-(pE+pI+pR)), E = n1*pE,I = n1*pI, R = n1*pR)

############# Running The Model ####################

resultsNV1 <- as.data.frame(ode(y=my_yinit, times=my_times, func=my_model,
	parms=my_pars, method='lsoda'))
############# Creating N ####################
resultsNV1$N <- (resultsNV1$S+resultsNV1$E+resultsNV1$I+resultsNV1$R)
############ Plotting Results ########################

colors <- c('Susceptible'='black','Exposed'='orange','Infected'='red', 'Recovered'='blue', 'Total'='pink')
ggplot(data=resultsNV1) + geom_line(aes(x=time,y=S,col='Susceptible')) + 
  geom_line(aes(x=time,y=E,col='Exposed')) +
	geom_line(aes(x=time,y=I,col='Infected')) + 
	geom_line(aes(x=time,y=R,col='Recovered')) +
  geom_line(aes(x=time,y=N,col='Total')) +
	scale_colour_manual(name="Legend",values=colors) +
	xlab('Year') + ylab('Number of People') +
  theme_bw() 

############ Setting Initial Conditions ###############

# initial starting values
my_yinit <- c(S = n2*(1-(pE+pI+pR)), E = n2*pE,I = n2*pI, R = n2*pR)

############# Running The Model ####################

resultsNV2 <- as.data.frame(ode(y=my_yinit, times=my_times, func=my_model,
	parms=my_pars, method='lsoda'))
############# Creating N ####################
resultsNV2$N <- (resultsNV2$S+resultsNV2$E+resultsNV2$I+resultsNV2$R)

############ Setting Initial Conditions ###############

# initial starting values
my_yinit <- c(S = n3*(1-(pE+pI+pR)), E = n3*pE,I = n3*pI, R = n3*pR)

############# Running The Model ####################

resultsNV3 <- as.data.frame(ode(y=my_yinit, times=my_times, func=my_model,
	parms=my_pars, method='lsoda'))
############# Creating N ####################
resultsNV3$N <- (resultsNV3$S+resultsNV3$E+resultsNV3$I+resultsNV3$R)

```

Vaccine:
```{r}
############# Model Equations ######################

my_model1 <- function(times,yinit,pars){
	
	with(as.list(c(yinit,pars)), {
	
		N = S + E + I + R
		dS <- w1*R + w2*V - beta*S
		dE <- (1-vc)*beta*S - lam*E
		dI <- lam*E - gamma*I
		dV <- vc*beta*S - w2*V
		dR <- gamma*I - w1*R

	return(list(c(dS, dE, dI, dV, dR)))})
	}

################## Setting Parameters ##################
#birth rate
b=0.00000575
#transmission parameter
beta=0.065585
#rate of becoming infectious
lam=0.07143
#rate of recovery
gamma=0.005 + (0.5*0.0165)
#rate of waning natural immunity
w1=0.01095
#vaccine coverage
vc=0.78
#rate of waning vaccine immunity (if different from natural)
w2= 0.12
#death rates
mu1=0.00005447
mu2=0.00005447
mu3=0.05
mu4=0.00005447
  
my_pars <- cbind(b,beta,lam,gamma,w1,vc,w2,mu1,mu2,mu3,mu4) 
  #cbind - combines data-frames, combines columns and rows respectively

############### Setting Time Frame ####################

start_time = 0 # start date (years)
end_time = 50 # end date (years)
my_times <- seq(start_time, end_time, by = 0.25) 
    # gives a sequence from start to end
											 # in increments of 1 month

############ Setting Initial Conditions ###############

# initial starting values
my_yinit <- c(S = n1*(1-(pE+pI+pR)), E = n1*pE,I = n1*pI, V=230462 ,R = n1*pR)

############# Running The Model ####################

results2 <- as.data.frame(ode(y=my_yinit, times=my_times, func=my_model1,
	parms=my_pars, method='lsoda'))
############# Creating N ####################
results2$N <- (results2$S+results2$E+results2$I+results2$V+results2$R)
############ Plotting Results ########################

colors <- c('Susceptible'='black','Exposed'='orange','Infected'='red','Vaccinated'='green','Recovered'='blue', 'Total'='pink')
ggplot(data=results2) + geom_line(aes(x=time,y=S,col='Susceptible')) + 
  geom_line(aes(x=time,y=E,col='Exposed')) +
	geom_line(aes(x=time,y=I,col='Infected')) + 
	geom_line(aes(x=time,y=V,col='Vaccinated')) +
	geom_line(aes(x=time,y=R,col='Recovered')) +
  geom_line(aes(x=time,y=N,col='Total')) +
	scale_colour_manual(name="Legend",values=colors) +
	xlab('Year') + ylab('Number of People') +
  theme_bw() 

```

Vaccine Expansion (new counties added in March 2023):
Note: for this model, the first few years of the un-expanded program should be included and then switched to greater population numbers and coverage change 
```{r}
my_model2 <- function(times,yinit,pars){
	
	with(as.list(c(yinit,pars)), {
	
		N = S + E + I + R
		dS <- w1*R + w2*V - beta*S
		dE <- (1-vc)*beta*S - lam*E
		dI <- lam*E - gamma*I
		dV <- vc*beta*S - w2*V
		dR <- gamma*I - w1*R

	return(list(c(dS, dE, dI, dV, dR)))})
	}

################## Setting Parameters ##################
#birth rate
b=0.00000575
#transmission parameter
beta=0.065585
#rate of becoming infectious
lam=0.07143
#rate of recovery
gamma=0.005 + (0.5*0.0165)
#rate of waning natural immunity
w1=0.01095
#vaccine coverage
vc=0.78
#rate of waning vaccine immunity (if different from natural)
w2= 0.12
#death rates
mu1=0.00005447
mu2=0.00005447
mu3=0.05
mu4=0.00005447
  
my_pars <- cbind(b,beta,lam,gamma,w1,vc,w2,mu1,mu2,mu3,mu4) 
  #cbind - combines data-frames, combines columns and rows respectively

############### Setting Time Frame ####################

start_time = 0 # start date (years)
end_time = 50 # end date (years)
my_times <- seq(start_time, end_time, by = 0.25) 
    # gives a sequence from start to end
											 # in increments of 1 month

############ Setting Initial Conditions ###############

# initial starting values
my_yinit <- c(S = n2*(1-(pE+pI+pR)), E = n2*pE,I = n2*pI, V=230462 ,R = n2*pR)

############# Running The Model ####################

results3 <- as.data.frame(ode(y=my_yinit, times=my_times, func=my_model2,
	parms=my_pars, method='lsoda'))
############# Creating N ####################
results3$N <- (results3$S+results3$E+results3$I+results3$V+results3$R)
############ Plotting Results ########################

colors <- c('Susceptible'='black','Exposed'='orange','Infected'='red','Vaccinated'='green','Recovered'='blue', 'Total'='pink')
ggplot(data=results3) + geom_line(aes(x=time,y=S,col='Susceptible')) + 
  geom_line(aes(x=time,y=E,col='Exposed')) +
	geom_line(aes(x=time,y=I,col='Infected')) + 
	geom_line(aes(x=time,y=V,col='Vaccinated')) +
	geom_line(aes(x=time,y=R,col='Recovered')) +
  geom_line(aes(x=time,y=N,col='Total')) +
	scale_colour_manual(name="Legend",values=colors) +
	xlab('Year') + ylab('Number of People') +
  theme_bw() 

```


All non-seasonal malaria regions
```{r}


my_model3 <- function(times,yinit,pars){
	
	with(as.list(c(yinit,pars)), {
	
		N = S + E + I + R
		dS <- w1*R + w2*V - beta*S
		dE <- (1-vc)*beta*S - lam*E
		dI <- lam*E - gamma*I
		dV <- vc*beta*S - w2*V
		dR <- gamma*I - w1*R

	return(list(c(dS, dE, dI, dV, dR)))})
	}

################## Setting Parameters ##################
#birth rate
b=0.00000575
#transmission parameter
beta=0.065585
#rate of becoming infectious
lam=0.07143
#rate of recovery
gamma=0.005 + (0.5*0.0165)
#rate of waning natural immunity
w1=0.01095
#vaccine coverage
vc=0.78
#rate of waning vaccine immunity (if different from natural)
w2= 0.12
#death rates
mu1=0.00005447
mu2=0.00005447
mu3=0.05
mu4=0.00005447
  
my_pars <- cbind(b,beta,lam,gamma,w1,vc,w2,mu1,mu2,mu3,mu4) 
  #cbind - combines data-frames, combines columns and rows respectively

############### Setting Time Frame ####################

start_time = 0 # start date (years)
end_time = 50 # end date (years)
my_times <- seq(start_time, end_time, by = 0.25) 
    # gives a sequence from start to end
											 # in increments of 1 month

############ Setting Initial Conditions ###############

# initial starting values
my_yinit <- c(S = n3*(1-(pE+pI+pR)), E = n3*pE,I = n3*pI, V=230462 ,R = n3*pR)

############# Running The Model ####################

results4 <- as.data.frame(ode(y=my_yinit, times=my_times, func=my_model3,
	parms=my_pars, method='lsoda'))
############# Creating N ####################
results4$N <- (results4$S+results4$E+results4$I+results4$V+results4$R)
############ Plotting Results ########################

colors <- c('Susceptible'='black','Exposed'='orange','Infected'='red','Vaccinated'='green','Recovered'='blue', 'Total'='pink')
ggplot(data=results4) + geom_line(aes(x=time,y=S,col='Susceptible')) + 
  geom_line(aes(x=time,y=E,col='Exposed')) +
	geom_line(aes(x=time,y=I,col='Infected')) + 
	geom_line(aes(x=time,y=V,col='Vaccinated')) +
	geom_line(aes(x=time,y=R,col='Recovered')) +
  geom_line(aes(x=time,y=N,col='Total')) +
	scale_colour_manual(name="Legend",values=colors) +
	xlab('Year') + ylab('Number of People') +
  theme_bw() 

```


# Cost-Effectiveness Analysis

```{r}
# Probabilities
pr_inpat = 0.008228  # probability of inpatient (hospitalization)
pr_outpat = 0.8375 # probability of outpatient treatment
pr_no_treat = 0.154272

pr_death_inpat = 0.0127
pr_death_outpat = 0.0063
pr_death_notreat = 0.0389


cases_NV1 <- sum(resultsNV1$I[1:20])*beta
cases_NV2 <- sum(resultsNV2$I[1:20])*beta
cases_NV3 <- sum(resultsNV3$I[1:20])*beta
cases_V1 <- sum(results2$I[1:20])*beta
cases_V2 <- sum(results3$I[1:20])*beta
cases_V3 <- sum(results4$I[1:20])*beta

cases_NV1
cases_NV2
cases_NV3
cases_V1
cases_V2
cases_V3

## Calculate the total number of cases, hospitalizations, and deaths
Hospitalizations_NV1 = cases_NV1*pr_inpat
Deaths_NV1= cases_NV1*(pr_inpat*pr_death_inpat +  
                              pr_outpat*pr_death_outpat + 
                              pr_no_treat*pr_death_notreat)
Hospitalizations_NV2 = cases_NV2*pr_inpat
Deaths_NV2= cases_NV2*(pr_inpat*pr_death_inpat +  
                              pr_outpat*pr_death_outpat + 
                              pr_no_treat*pr_death_notreat)
Hospitalizations_NV3 = cases_NV3*pr_inpat
Deaths_NV3= cases_NV3*(pr_inpat*pr_death_inpat +  
                              pr_outpat*pr_death_outpat + 
                              pr_no_treat*pr_death_notreat)
Hospitalizations_V1 = cases_V1*pr_inpat
Deaths_V1= cases_V1*(pr_inpat*pr_death_inpat +  
                              pr_outpat*pr_death_outpat + 
                              pr_no_treat*pr_death_notreat)
Hospitalizations_V2 = cases_V2*pr_inpat
Deaths_V2= cases_V2*(pr_inpat*pr_death_inpat +  
                              pr_outpat*pr_death_outpat + 
                              pr_no_treat*pr_death_notreat)
Hospitalizations_V3 = cases_V3*pr_inpat
Deaths_V3= cases_V3*(pr_inpat*pr_death_inpat +  
                              pr_outpat*pr_death_outpat + 
                              pr_no_treat*pr_death_notreat)

# Cases averted with vaccine:
CasesAverted_0 = cases_NV1 - cases_V1
# Hospitalizations averted:
HospAverted_0 = Hospitalizations_NV1 - Hospitalizations_V1
# Deaths Averted
DeathsAverted_0 = Deaths_NV1- Deaths_V1

# Cases averted with vaccine:
CasesAverted_E = cases_NV2 - cases_V2
# Hospitalizations averted:
HospAverted_E = Hospitalizations_NV2 - Hospitalizations_V2
# Deaths Averted
DeathsAverted_E = Deaths_NV2 - Deaths_V2

# Cases averted with vaccine:
CasesAverted_T = cases_NV3 - cases_V3
# Hospitalizations averted:
HospAverted_T = Hospitalizations_NV3 - Hospitalizations_V3
# Deaths Averted
DeathsAverted_T = Deaths_NV3 - Deaths_V3


print(CasesAverted_0)
print(HospAverted_0)
print(DeathsAverted_0)

print(CasesAverted_E)
print(HospAverted_E)
print(DeathsAverted_E)

print(CasesAverted_T)
print(HospAverted_T)
print(DeathsAverted_T)

############ Question 8 ##############

# Calculate number of people vaccinated in each scenario
# use same runs as in question 4

# Routine: infants are vaccinated as they age from age class 1 to age class 2
# vacc doses = sum of (aging rate from age class 1 to class 2 * vax coverage * (S + S_R + R + C + I_A)),
# with the last two states representing "wasted" doses
cov = 0.78
Doses_V1 = sum(cov*sum(results2$S[1:12]))+230462
Doses_V2 = sum(cov*sum(results3$S[1:20]))+230462
Doses_V3 = sum(cov*sum(results4$S[1:20]))+230462
Doses_V1 
Doses_V2 
Doses_V3 


############ Question 9 ##############

# Calculate ICERs for different strategy pairs

# Routine vaccination vs no vaccination
# Calculate Cost of Routine - Cost of No vaccination = net cost
# Calculate DALYs with no vaccination -DALYs with Routine vaccination = DALYs averted (net benefit)

# Costs
cost_V = 3.01 + 2.46  # routine vaccine 
cost_in_trt = 58  # inpatient treatment
cost_out_trt = 2.50 # outpatient treatment

# DALY info 
daly_wt = 0.205
dur_inf = ((0.845728*14)+((1-0.845728)*121))/365
life_exp = 66.1
 #life expectancy at death for each age group


### Costs of each strategy ###

# Cost of No vacc = 0 vaccine costs + cost of inpatient treatment * number of hospitalizations + cost of outpatient treatment * number outpatient treated
Cost_NV1 = cost_in_trt*Hospitalizations_NV1 + 
  cost_out_trt*cases_NV1*pr_outpat 
Cost_NV2 = cost_in_trt*Hospitalizations_NV2 + 
  cost_out_trt*cases_NV2*pr_outpat 
Cost_NV3 = cost_in_trt*Hospitalizations_NV3 + 
  cost_out_trt*cases_NV3*pr_outpat 

# Cost of Routine = Cost of Routine Vaccination * number of vaccinations + 
# cost of inpatient treatment * number of hospitalizations + cost of outpatient treatment * number outpatient treated

Cost_V1 = cost_V*Doses_V1 + 
  cost_in_trt*Hospitalizations_V1 + 
  cost_out_trt*pr_outpat*cases_V1
Cost_V2 = cost_V*Doses_V2 + 
  cost_in_trt*Hospitalizations_V2 + 
  cost_out_trt*pr_outpat*cases_V2
Cost_V3 = cost_V*Doses_V3 + 
  cost_in_trt*Hospitalizations_V3 + 
  cost_out_trt*pr_outpat*cases_V3

print(paste0("Cost of Routine Vaccination Scenario 1 ", round(Cost_V1,0)))
print(paste0("Cost of No Vaccination Scenario 1 ", round(Cost_NV1,0)))
print(paste0("Cost of Routine Vaccination Scenario 2 ", round(Cost_V2,0)))
print(paste0("Cost of No Vaccination Scenario 2 ", round(Cost_NV2,0)))
print(paste0("Cost of Routine Vaccination Scenario 3 ", round(Cost_V3,0)))
print(paste0("Cost of No Vaccination Scenario 3 ", round(Cost_NV3,0)))



# DALY = YLD + YLL
# YLD (years lived with disability ) = Number of cases x duration of infection x disability weight
# YLL (years of life lost) = number of deaths*life expectancy at age of death
# # means we need to calculate separately for each age group

# DALYs for No Vaccination    
DALYs_NV1= cases_NV1*dur_inf*daly_wt + 
  (pr_inpat*pr_death_inpat + pr_outpat*pr_death_outpat+pr_no_treat*pr_death_notreat)*
  (sum(resultsNV1$I[1:20]))*(life_exp)
DALYs_NV2= cases_NV2*dur_inf*daly_wt + 
  (pr_inpat*pr_death_inpat + pr_outpat*pr_death_outpat+pr_no_treat*pr_death_notreat)*
  (sum(resultsNV2$I[1:20]))*(life_exp)
DALYs_NV3= cases_NV3*dur_inf*daly_wt + 
  (pr_inpat*pr_death_inpat + pr_outpat*pr_death_outpat+pr_no_treat*pr_death_notreat)*
  (sum(resultsNV3$I[1:20]))*(life_exp)

# DALYs for Vaccine
DALYs_V1=
  cases_V1*dur_inf*daly_wt + 
  (pr_inpat*pr_death_inpat + pr_outpat*pr_death_outpat+pr_no_treat*pr_death_notreat)*
  (sum(results2$I[1:20]))*(life_exp)
DALYs_V2=
  cases_V2*dur_inf*daly_wt + 
  (pr_inpat*pr_death_inpat + pr_outpat*pr_death_outpat+pr_no_treat*pr_death_notreat)*
  (sum(results3$I[1:20]))*(life_exp)
DALYs_V3=
  cases_V3*dur_inf*daly_wt + 
  (pr_inpat*pr_death_inpat + pr_outpat*pr_death_outpat+pr_no_treat*pr_death_notreat)*
  (sum(results4$I[1:20]))*(life_exp)

V1_NV1_ICER = (Cost_V1-Cost_NV1)/ (DALYs_NV1 - DALYs_V1)
print(V1_NV1_ICER)

V2_NV2_ICER = (Cost_V2-Cost_NV2)/ (DALYs_NV2 - DALYs_V2)
print(V2_NV2_ICER)

V3_NV3_ICER = (Cost_V3-Cost_NV3)/ (DALYs_NV3 - DALYs_V3)
print(V3_NV3_ICER)

costsV <- c(Cost_V1,Cost_V2,Cost_V3)
costsNV <- c(Cost_NV1,Cost_NV2,Cost_NV3)
DALYsV <- c(DALYs_V1,DALYs_V2,DALYs_V3)
DALYsNV <- c(DALYs_NV1,DALYs_NV2,DALYs_NV3)
plot((costsV-costsNV),(DALYsNV-DALYsV), xlab="DALYs Averted",ylab="DALYs Averted")

```


